Containers and R: the Rockerverse and beyond

### 2019-10-18 manuscript.Rmd
Added draft of holepunch section

---
title: "Containers and R: the Rockerverse and beyond"
author:
  - name: Daniel Nüst
    affiliation: Institute for Geoinformatics, University of Münster
    address:
    - Heisenbergstr. 2, 48149 Münster, Germany
    email: daniel.nuest@uni-muenster.de
  - name: Carl Boettiger
    #affiliation: -
    #email: -
  - name: Robrecht Cannoodt
    #affiliatin: -
    #email: -
  - name: Dirk Eddelbuettel
    affiliation: University of Illinois at Urbana-Champaign
    email: dirkd@eddelbuettel.com
  - name: Mark Edmondson
    affiliation: IIH Nordic A/S, Google Developer Expert for GCP
    email: mark@markedmondson.me
  - name: Colin Fay
    affiliation: ThinkR
    email: contact@colinfay.me
  - name: Ben Marwick
    #affiliation: -
    #email: -
  - name: Karthik Ram
    #affiliation: -
    #email: -
  - name: Noam Ross
    #affiliation: -
    #email: -
  - name: Nan Xiao
    affiliation: Seven Bridges Genomics
    address: "529 Main St, Suite 6610 \\newline Charlestown, MA 02129, USA \\newline ORCiD: 0000-0002-0250-5673"
    email: me@nanx.me
  - name: Lori Shepherd
    affiliation: Roswell Park Comprehensive Cancer Center
    address:
    - Elm & Carlton Streets, Buffalo, New York, 14263 USA
    email: lori.shepherd@roswellpark.org
  - name: Nitesh Turaga
    affiliation: Roswell Park Comprehensive Cancer Center
    address:
    - Elm & Carlton Streets, Buffalo, New York, 14263 USA
    email: nitesh.turaga@roswellpark.org
abstract: >
  The Rocker project provides widely-used Docker images for R across different application scenarios.
  This articles surveys describe downstream projects building upon Rocker.
  We also look beyond Rocker to other projects connecting containerisation with R.
  These use cases and the diversity of applications demonstrate the power of Rocker and containerisation for collaboration, effectivity, scalability, and transparency.
# blank footnote? https://tex.stackexchange.com/questions/170511/footnotes-without-numbering
preamble: >
  \usepackage{longtable}
  \usepackage{hyperref}
output: rticles::rjournal_article
#  includes:
#    in_header: preamble.tex
---

@@

- Binder
  - [`holepunch`](https://github.com/karthik/holepunch) streamlines making an R research compendium Binder-ready [@karthik]
  - [`repo2docker`](https://repo2docker.readthedocs.io/en/latest/config_files.html) (`install.R`, `DESCRPTION`, `runtime.txt`) based on the Rocker images [@nuest]
- GPU [@noamross, @cboettig]
- Gigantum stack
- in education (ready to use teaching/learning environments)
  - [`r-db`](https://github.com/ColinFay/r-db) [@ColinFay]
- [RCloud](https://github.com/att/rcloud/tree/master/docker) ?

@@

DN conceived of the presented idea and \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}.
CB ..
RC ..
DE ..
ME ..
CF wrote the section about `r-online`, `dockerfiler`, `r-ci` and `r-db`.
BW ..
KR ..
NR ..
NX wrote the section on liftr.
LS \& NT wrote the section on Bioconductor.
All authors contributed to the discussion and outlook section and approved the final version.
This articles was collaboratively written at \href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
The \href{https://github.com/nuest/rockerverse-paper/graphs/contributors}{contributors page} and \href{https://github.com/nuest/rockerverse-paper/commits/master}{commit history} provide a detailed view on the respective contributions.

>>>>

---
title: "Containers and R: the Rockerverse and beyond"
author:
  - name: Daniel Nüst
    affiliation: Institute for Geoinformatics, University of Münster
    address:
    - Heisenbergstr. 2, 48149 Münster, Germany
    email: daniel.nuest@uni-muenster.de
  - name: Carl Boettiger
    #affiliation: -
    #email: -
  - name: Robrecht Cannoodt
    #affiliatin: -
    #email: -
  - name: Dirk Eddelbuettel
    affiliation: University of Illinois at Urbana-Champaign
    email: dirkd@eddelbuettel.com
  - name: Mark Edmondson
    affiliation: IIH Nordic A/S, Google Developer Expert for GCP
    email: mark@markedmondson.me
  - name: Colin Fay
    affiliation: ThinkR
    email: contact@colinfay.me
  - name: Ben Marwick
    #affiliation: -
    #email: -
  - name: Karthik Ram
    affiliation: Berkeley Institute for Data Science, University of California, Berkeley, CA 94720 USA
    email: karthik.ram@berkeley.edu
  - name: Noam Ross
    #affiliation: -
    #email: -
  - name: Nan Xiao
    affiliation: Seven Bridges Genomics
    address: "529 Main St, Suite 6610 \\newline Charlestown, MA 02129, USA \\newline ORCiD: 0000-0002-0250-5673"
    email: me@nanx.me
  - name: Lori Shepherd
    affiliation: Roswell Park Comprehensive Cancer Center
    address:
    - Elm & Carlton Streets, Buffalo, New York, 14263 USA
    email: lori.shepherd@roswellpark.org
  - name: Nitesh Turaga
    affiliation: Roswell Park Comprehensive Cancer Center
    address:
    - Elm & Carlton Streets, Buffalo, New York, 14263 USA
    email: nitesh.turaga@roswellpark.org
abstract: >
  The Rocker project provides widely-used Docker images for R across different application scenarios.
  This articles surveys describe downstream projects building upon Rocker.
  We also look beyond Rocker to other projects connecting containerisation with R.
  These use cases and the diversity of applications demonstrate the power of Rocker and containerisation for collaboration, effectivity, scalability, and transparency.
# blank footnote? https://tex.stackexchange.com/questions/170511/footnotes-without-numbering
preamble: >
  \usepackage{longtable}
  \usepackage{hyperref}
output: rticles::rjournal_article
#  includes:
#    in_header: preamble.tex
---

@@

- Binder

The [Binder project](https://mybinder.readthedocs.io/en/latest/), maintained by the team behind Jupyter, makes it possible for users to create and share their computing environments with others. A deployed version of this service called binder hub allows anyone with access to a web browser and an internet connection to launch a temporary instance of these custom environments and execute any notebooks contained within. From a reproducibility standpoint, Binder makes it exceedingly easy to compile a paper, visualize data, and run small examples from papers or tutorials without the need for any local installation.

To set up Binder for a project, a user typically starts at an instance of a Binderhub (examples include mybinder.org and binder.pangeo.io) and passes the location of a hosted Git repository. Using internal tools such as [repo2docker](https://repo2docker.readthedocs.io/en/latest/config_files.html), the binderhub builds a Dockerfile by parsing the contents of all code contained within. While this approach works well for most run of the mill Python projects, it is not so seamless for R projects. For any R projects that use the Tidyverse suite [citation in review], the time and resources required to build all dependencies from source can often time out before completion, making it frustrating for the average R user.

[`holepunch`](https://github.com/karthik/holepunch) is a R package that was designed to remove some of these limitations and make Binder more accessible to novice R users. `holepunch` achieves this by leveraging Rocker images that contains the Tidyverse along special Jupyter dependencies, and only installs additional packages from CRAN and Bioconductor that are not already part of these images. Starting with the Binder/Tidyverse base images eliminates a large part of the build time and in most cases results in a binder instance launching within a minute. `holepunch` as a side effect also creates a DESCRIPTION file which then turns any project into a research compendium [Marwick et al]. The Dockerfile included with the project can also be used to launch a RStudio server locally independent of binder which is especially useful when more computational resources are required.

  - [`repo2docker`](https://repo2docker.readthedocs.io/en/latest/config_files.html) (`install.R`, `DESCRPTION`, `runtime.txt`) based on the Rocker images [@nuest]
- GPU [@noamross, @cboettig]
- Gigantum stack
- in education (ready to use teaching/learning environments)
  - [`r-db`](https://github.com/ColinFay/r-db) [@ColinFay]
- [RCloud](https://github.com/att/rcloud/tree/master/docker) ?

@@

DN conceived of the presented idea and \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}.
CB ..
RC ..
DE ..
ME ..
CF ..
BW ..
KR ..
NR ..
NX wrote the section on liftr.
LS \& NT wrote the section on Bioconductor.
All authors contributed to the discussion and outlook section and approved the final version.
This articles was collaboratively written at \href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
The \href{https://github.com/nuest/rockerverse-paper/graphs/contributors}{contributors page} and \href{https://github.com/nuest/rockerverse-paper/commits/master}{commit history} provide a detailed view on the respective contributions.

### 2019-10-18 manuscript.tex
Added draft of holepunch section

\emph{Renjin} is an interpreter for the R language running on top of the
\href{https://en.wikipedia.org/wiki/Java_virtual_machine}{Java Virtual
Machine} (JVM) providing full two-way access between Java and R code
\citep{wikipedia_renjin_2018}. It was developed to combine the benefits
of R, such scripting and extension packages, with the JVM's advantages
in the areas of security, cross-platform availability, and established
position in enterprise settings. R extension packages need to be
specially compiled and are distributed via the Java package manager
\href{https://en.wikipedia.org/wiki/Apache_Maven}{Apache Maven}, cf.
\url{http://packages.renjin.org/packages} for available packages.
Packages are loaded on demand, i.e.~at the first call to
\texttt{library()}. Not all R packages, especially one linking to binary
libraries, are available, e.g. \texttt{rgdal}
\footnote{\href{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}}.

@@

  \begin{itemize}
  \tightlist
  \item
    I don't think \texttt{docker\_run} is main focus for containers for
    this package, its more how it uses Docker to launch custom
    environments that can be templated (e.g.~rstudio, shiny etc.) and
    for parrallisation e.g. \texttt{library(future)} using GCE VMs as
    endpoints. {[}@MarkEdmondson1234{]}
  \end{itemize}
\item
  \href{https://github.com/sckott/analogsea}{\texttt{analogsea}}
\item
  \href{https://github.com/wch/harbor/}{\texttt{harbor}}
\item
  \href{https://github.com/cboettig/dockermachine}{\texttt{dockermachine}}
\end{itemize}

@@

  \begin{itemize}
  \tightlist
  \item
    the most popular functionality for \texttt{googleComputeEngineR} is
    its use of Docker to enable parallel processing across VMS -
    \href{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}{some
    demos here}
  \end{itemize}
\item
  Google Cloud Run - CaaS (Containers as a Service) that lets you launch
  a Docker container without worrying about underlying infrastructure.
  An R implementation is shown here at
  \href{https://github.com/MarkEdmondson1234/cloudRunR}{cloudRunR} which
  uses it to create a scalable R plumber API.
\item
  \href{https://www.rplumber.io/docs/hosting.html\#docker}{\texttt{plumber}}
\item
  \texttt{batchtools} \citep{Lang2017batchtools} can
  \href{https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html}{schedule
  jobs with Docker Swarm}
\item
  scalable deployments, e.g.~start with numerous Shiny talks mentioning
  Rocker at useR!2017
\item
  \href{https://github.com/dynverse/dynmethods}{dynmethods}
  {[}@rcannood{]}: In order to evaluate ±50 computational methods which
  all used different environments (R, Python, C++, \ldots{}), we wrapped
  each of them in a docker container and can execute these methods from
  R. Again, all of these containers are being built on travis-ci, and
  will only be pushed to docker hub if the integration test succeeds.
\item
  \href{https://www.shinyproxy.io/}{ShinyProxy}
  \href{https://github.com/openanalytics/shinyproxy/blob/master/src/main/java/eu/openanalytics/services/DockerService.java\#L388}{creates
  a container} for each user
\end{itemize}

@@

\address{%
Karthik Ram\\
\\
\\
}

>>>>

\emph{Renjin} is an interpreter for the R language running on top of the
\href{https://en.wikipedia.org/wiki/Java_virtual_machine}{Java Virtual
Machine} (JVM) providing full two-way access between Java and R code
\citep{wikipedia_renjin_2018}. It was developed to combine the benefits
of R, such scripting and extension packages, with the JVM's advantages
in the areas of security, cross-platform availability, and established
position in enterprise settings. R extension packages need to be
specially compiled and are distributed via the Java package manager
\href{https://en.wikipedia.org/wiki/Apache_Maven}{Apache Maven},
cf.~\url{http://packages.renjin.org/packages} for available packages.
Packages are loaded on demand, i.e.~at the first call to
\texttt{library()}. Not all R packages, especially one linking to binary
libraries, are available, e.g.~\texttt{rgdal}
\footnote{\href{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}}.

@@

  \begin{itemize}
  \tightlist
  \item
    I don't think \texttt{docker\_run} is main focus for containers for
    this package, its more how it uses Docker to launch custom
    environments that can be templated (e.g.~rstudio, shiny etc.) and
    for parrallisation e.g.~\texttt{library(future)} using GCE VMs as
    endpoints. {[}@MarkEdmondson1234{]}
  \end{itemize}
\item
  \href{https://github.com/sckott/analogsea}{\texttt{analogsea}}
\item
  \href{https://github.com/wch/harbor/}{\texttt{harbor}}
\item
  \href{https://github.com/cboettig/dockermachine}{\texttt{dockermachine}}
\end{itemize}

@@

  \begin{itemize}
  \tightlist
  \item
    the most popular functionality for \texttt{googleComputeEngineR} is
    its use of Docker to enable parallel processing across VMS -
    \href{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}{some
    demos here}
  \end{itemize}
\item
  Google Cloud Run - CaaS (Containers as a Service) that lets you launch
  a Docker container without worrying about underlying infrastructure.
  An R implementation is shown here at
  \href{https://github.com/MarkEdmondson1234/cloudRunR}{cloudRunR} which
  uses it to create a scalable R plumber API.
\item
  \href{https://www.rplumber.io/docs/hosting.html\#docker}{\texttt{plumber}}
\item
  \texttt{batchtools} \citep{Lang2017batchtools} can
  \href{https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html}{schedule
  jobs with Docker Swarm}
\item
  scalable deployments, e.g.~start with numerous Shiny talks mentioning
  Rocker at useR!2017
\item
  \href{https://github.com/dynverse/dynmethods}{dynmethods}
  {[}@rcannood{]}: In order to evaluate ±50 computational methods which
  all used different environments (R, Python, C++, \ldots), we wrapped
  each of them in a docker container and can execute these methods from
  R. Again, all of these containers are being built on travis-ci, and
  will only be pushed to docker hub if the integration test succeeds.
\item
  \href{https://www.shinyproxy.io/}{ShinyProxy}
  \href{https://github.com/openanalytics/shinyproxy/blob/master/src/main/java/eu/openanalytics/services/DockerService.java\#L388}{creates
  a container} for each user
\end{itemize}

@@

\address{%
Karthik Ram\\
Berkeley Institute for Data Science, University of California, Berkeley,
CA 94720 USA\\
\\
}
\href{mailto:karthik.ram@berkeley.edu}{\nolinkurl{karthik.ram@berkeley.edu}}
